A weighted sampling algorithm to designing RNA sequences with target secondary structure and nucleotide distribution

### 2013-01-08 ISMB/methods_ISMB.tex
fix typos

We introduce a probabilistic model for the design of RNA sequences with a specific $\Gb+\Cb$ content and folding into a predefined secondary structure.
For the sake of simplicity, we choose to base this proof-of-concept implementation on a simplified free-energy function $\ES(\cdot)$, which only considers the contributions of 
stacked canonical base-pairs. We show how a modification of the dynamic programming scheme used in the \RNAmutants allows for the sampling of good and diverse design candidates, in linear time and space complexities.

@@

\begin{figure}
\resizebox{\textwidth}{!}{\input{FigStochasticBacktrack.pgf}}
\caption{Three main cases in the stochastic backtrack procedure over an interval $[i,j]$: Either position $i$ is left unpaired (top), a base-pair $(i,j)$ is formed, stacking onto $(i+1,j-1)$ (middle), or $i$ is paired somewhere else, defining two regions on which subsequent recursive calls are needed (bottom). Positions indicated in red are assigned during at this stage of the backtrack.\label{fig:stochastic}}
\end{figure}

>>>>

We introduce a probabilistic model for the design of RNA sequences with a specific \GCContent and folding into a predefined secondary structure.
For the sake of simplicity, we choose to base this proof-of-concept implementation on a simplified free-energy function $\ES(\cdot)$, which only considers the contributions of 
stacked canonical base-pairs. We show how a modification of the dynamic programming scheme used in the \RNAmutants allows for the sampling of good and diverse design candidates, in linear time and space complexities.

@@

\begin{figure}
\resizebox{\textwidth}{!}{\input{FigStochasticBacktrack.pgf}}
\caption{Three main cases in the stochastic backtrack procedure over an interval $[i,j]$: Either position $i$ is left unpaired (top), a base-pair $(i,j)$ is formed, stacking onto $(i+1,j-1)$ (middle), or $i$ is paired somewhere else, defining two regions on which subsequent recursive calls are needed (bottom). Positions indicated in red are assigned during this stage of the backtrack.\label{fig:stochastic}}
\end{figure}

### 2013-01-08 ISMB/results_ISMB.tex
fix typos

We present the average times spent by \texttt{Incarnation} and \RNAinverse to generate one sequence
with the required \texttt{C+G} content in Figure~\ref{fig:time}. As expected, the time grows linearly
in function of the length of the structures for \texttt{Incarnation}.  A highly interesting feature
is the huge decrease in time of computation when there is $15\%$ free 
nucleotides inside the loops of the secondary structure.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/time_rnastrand_clustered_rnainverse_100samples_fix}
	\caption{Average time in seconds to generate one sequence. We explicitly show 
	the time spent by \texttt{Incarnation} (full line) and \texttt{RNAinverse} (dashed) for various \texttt{C+G} contents. The first plot is in function
	of the length of the structures. The second is in function of the stack
	density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	the proportion of free nucleotides (i.e. not base paired) inside loops.}
	\label{fig:time}	
\end{figure}

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \texttt{Incarnation} in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \texttt{Incarnation} sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is not a good predictor for the quality of the results. Instead,
a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
On top of that, with a \texttt{C+G} content of $50\%$ and more, at least
one solutions was found for almost all structures. The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last subfigure. Even with a high \texttt{C+G} content a our method was not able to generate solutions for a few
 structures.

@@

To evaluate the global quality of \texttt{Incarnation} sequences, we show
in Figure~\ref{fig:ss_sens} the ratio of well predicted base pairs in the
MFE structure of our sampled sequences. We can observe that, in all cases,
the hardest sequence to design are those with an extremely low \texttt{C+G}
content. As anticipated, those are the sequences with the weakest bonds.
We notice that the most accurate sequences yield a \texttt{C+G} content
of $70\pm 10\%$.

@@

A desirable feature in sequence design, is to produce samples with a high
diversity. We show in Figure~\ref{fig:nb_sols_entropy} the number of correct
solutions as their average entropy and base pair entropy, since 
\texttt{Incarnation}  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \texttt{C+G} content and the percentage of free nucleotides. Our method is able to generate

Also, a critical properties in RNA sequence design is 
the frequency of the MFE. 
The sequence should be in its target conformation long enough to
perform the desired action. Presented in Figure~\ref{fig:freq} we see that
there is a slow decline of the frequency with the increase in size. Yet,
for an average \texttt{C+G} content, the frequency stays over $10\%$ even
at size of $100$ nucleotides.

>>>>

We present the average times spent by \texttt{Incarnation} and \RNAinverse to generate one sequence
with the required \GCContent in Figure~\ref{fig:time}. As expected, the time grows linearly
in function of the length of the structures for \texttt{Incarnation}.  A highly interesting feature
is the huge decrease in time of computation when there is $15\%$ free 
nucleotides inside the loops of the secondary structure.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/time_rnastrand_clustered_rnainverse_100samples_fix}
	\caption{Average time in seconds to generate one sequence. We explicitly show 
	the time spent by \texttt{Incarnation} (full line) and \texttt{RNAinverse} (dashed) for various \GCContent. The first plot is in function
	of the length of the structures. The second is in function of the stack
	density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	the proportion of free nucleotides (i.e. not base paired) inside loops.}
	\label{fig:time}	
\end{figure}

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \texttt{Incarnation} in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \texttt{Incarnation} sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is not a good predictor for the quality of the results. Instead,
a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
 The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last column of Figure~\ref{fig:mfe_struct_solved}.
Even with a \GCContent of $50\%$ and more, when at least
one solutions was found for almost all structures, most of the samples 
do not fold into the target. A few structures remained unsolved under
all \GCContent.

@@

To evaluate the global quality of \texttt{Incarnation} sequences, we show
in Figure~\ref{fig:ss_sens} the ratio of well predicted base pairs in the
MFE structure of our sampled sequences. We can observe that, in all cases,
the hardest sequence to design are those with an extremely low \GCContent. As anticipated, those are the sequences with the weakest bonds.
We notice that the most accurate sequences yield a \GCContent
of $70\pm 10\%$.

@@

A desirable feature in sequence design, is to produce samples with a high
diversity. We show in Figure~\ref{fig:nb_sols_entropy} the number of correct
solutions as their average entropy and base pair entropy, since 
\texttt{Incarnation}  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \GCContent and the percentage of free nucleotides. Our method is able to generate

Also, a critical properties in RNA sequence design is 
the frequency of the MFE. 
The sequence should be in its target conformation long enough to
perform the desired action. Presented in Figure~\ref{fig:freq} we see that
there is a slow decline of the frequency with the increase in size. Yet,
for an average \GCContent, the frequency stays over $10\%$ even
at size of $100$ nucleotides.
